Is high social class always beneficial for survival? A study of northern Sweden 1801--2013 \footnote{Short; "Is high social class beneficial for survival?"}

### 2018-10-11 future.Rmd
error tracking

```{r readalldata, eval = FALSE}
source("R/get.file.R")
##skume <- get.file("skume")
load("data/skume.rda")
skume <- skume[order(skume$id, skume$enter), ]
skume$id <- paste("pop", skume$id, sep = "_")
##
skume1 <- age.window(skume, c(40, 65))  ###### OJOJOJ!!!!!!!!!!!!!!!!!!!!!!!!!!! (40, 65)
skume1 <- oneSpell(skume1)
##
skume2 <- age.window(skume, c(65, 100))
skume2 <- oneSpell(skume2)
##################################################
##foli <- get.file("foli")
load("data/foli.rda")
foli$civst[is.na(foli$civst)] <- "unmarried"
foli <- foli[order(foli$id, foli$enter), ]
foli$AR <- foli$ULORSAK <- NULL
##
foli1 <- age.window(foli, c(40, 65))
foli1 <- oneSpell(foli1)
##
foli2 <- age.window(foli, c(65, 100))
foli2 <- oneSpell(foli2)
##################################################
source("CreData/R/addPer.R")
skume1 <- addPer(skume1, cuts = seq(1801, 1951, by = 25))
skume2 <- addPer(skume2, cuts = seq(1801, 1951, by = 25))
##
foli1 <- addPer(foli1, cuts = c(1976, 1986, 1996, 2005, 2014))
foli2 <- addPer(foli2, cuts = c(1976, 1986, 1996, 2005, 2014))
##################################################
vb1 <- rbind(skume1, foli1)
vb1$period <- factor(vb1$period)
vb1Tab <- toTpch(vb1[, !(colnames(vb1) %in% c("id", "birthdate"))], cuts = seq(40, 65, by = 1))
##
vb2 <- rbind(skume2, foli2)
vb2$period <- factor(vb2$period)
vb2Tab <- toTpch(vb2[, !(colnames(vb2) %in% c("id", "birthdate"))], cuts = seq(65, 100, by = 1))
vb <- rbind(vb1, vb2)
vb$exposure <- vb$exit - vb$enter # For statistics and figures purposes
vbTab <- rbind(vb1Tab, vb2Tab) # NEW
```

```{r nydata}
vbTab <- readRDS("data/vbTab.rds")
vb1Tab <- readRDS("data/vb1Tab.rds")
vb2Tab <- readRDS("data/vb2Tab.rds")
```

@@

```{r hazcalw}
source("R/getWeights.R")
labb <- levels(vbTab$period)
nper <- length(labb)
his <- levels(vb1Tab$hisclass)
nhis <- length(his)
##
resm <- matrix(0, nrow = nhis, ncol = nper)
pdm <-  matrix(0, nrow = nhis, ncol = nper)
for (i in 1:nper){
    dat <- vb1Tab[vb1Tab$sex == "male" & vb1Tab$period == labb[i], ]
    ##dat <- vb1[vb1$sex == "male" & vb1$period == labb[i], ]
    ##dat$wght <- getWeights(dat)
    ##fit <- coxph(Surv(enter, exit, event) ~ strata(hisclass), weights = wght, data = dat)
    ##x <- survfit(fit)
    for (j in seq_along(his)){
        fit <- tpchreg(event ~ offset(log(exposure)) + age, data = dat[dat$hisclass == his[j], ], cuts = seq(40, 65, by = 1))
        haz <- c(0, fit$hazards, fit$hazards[length(fit$hazards)])
        pdm[j, i] <- ppch(65, cuts = fit$cuts, levels = haz)
    }
    ##resm[, i] <- summary(x, rmean = 65)$table[, "*rmean"]
    ##pdm[, i] <- 1 - x$surv[cumsum(x$strata)]
}
##
resw <- matrix(0, nrow = nhis, ncol = nper)
pdw <-  matrix(0, nrow = nhis, ncol = nper)
for (i in 1:nper){
    dat <- vb1Tab[vb1Tab$sex == "female" & vb1Tab$period == labb[i], ]
    ##dat <- vb1[vb1$sex == "female" & vb1$period == labb[i], ]
    ##dat$wght <- getWeights(dat)
    ##fit <- coxph(Surv(enter, exit, event) ~ strata(hisclass), weights = wght, data = dat)
    ##x <- survfit(fit)
    for (j in seq_along(his)){
        fit <- tpchreg(event ~ offset(log(exposure)) + age, data = dat[dat$hisclass == his[j], ], cuts = seq(40, 65, by = 1))
        haz <- c(0, fit$hazards, fit$hazards[length(fit$hazards)])
        pdw[j, i] <- ppch(65, cuts = fit$cuts, levels = haz)
    }
    ##resw[, i] <- summary(x, rmean = 65)$table[, "*rmean"]
    ##pdw[, i] <- 1 - x$surv[cumsum(x$strata)]
}

@@

<!--
```{r cumsw, fig.cap = "Differences in expected years of survival until the age of 65, given survival to age 40, by period and hisclass. Reference is the working class.", fig.scap = "Differences in expected years of life at age 40", fig.height = 8, out.height = "14cm"}
source("R/plotED.glm.R")
oldpar <- par(mfrow = c(2, 1))
midper <- c(1812.5, 1837.5, 1862.5, 1887.5, 1912.5, 1937.5, 1981, 1991, 2000.5, 2009.5)
refclw <- resw[3, ]
plot(midper, resw[1, ] - refclw, col = scol[1], type = "b", ylim = c(-3, 3), xlab = "Period", xlim = c(1810, 2015),
     ylab = "Years", axes = FALSE, main = "Women, 40-64", lty = 1, pch = 1)

axis(1, at = midper, labels = labb, las = 2, cex.axis = 0.5)
axis(2, las = 1)
axis(4, las = 1)
abline(h = 65)
box()
for (i in 2:5){
    lines(midper, resw[i, ] - refclw, col = scol[i], type = "b", pch = i, lty = i)
}
legend("topright", legend = his, col = scol, lty = 1:5, pch = 1:5, cex = 0.7, bty = "n")
refclm <- resm[3, ]
plot(midper, resm[1, ] - refclm, col = scol[1], type = "b", ylim = c(-3, 3), xlab = "Period",
     ylab = "Years", axes = FALSE, main = "Men, 40-64", lty = 1, pch = 1, xlim = c(1810, 2015))
axis(1, at = midper, labels = labb, las = 2, cex.axis = 0.5)
axis(2, las = 1)
axis(4, las = 1)
abline(h = 65)
box()
for (i in 2:5){
    lines(midper, resm[i, ] - refclm, col = scol[i], type = "b", lty = i, pch = i)
}
legend("topright", legend = his, col = scol, lty = 1:5, pch = 1:5, cex = 0.7, bty = "n")
##woutp <- plotED.glm(resw, col = scol, logScale = TRUE, ylab = "H(65|30)",
  ##              main = "Women 30-64", ylim = c(0, 1))##, exclude = 5:6)
##moutp <- plotED.glm(resm, col = scol, logScale = TRUE, ylab = "H(65|30)",
  ##              main = "Men 30-64", ylim = c(0, 1))##, exclude = 5:6)
par(oldpar)
```

-->
Differences in mortality in the age group 40--64 are getting very small as we move forward in time: Almost all forty-year-olds today survive beyond 65, regardless of social class. In Sweden of the year 2014, the probability of surviving age 65 for a forty-year-old man is 91 per cent  (a period measure). The corresponding figure for a forty-year-old woman is 94 per cent. A relevant measure of comparison is the probability (risk) of dying before age 65 for a forty-year-old, remembering that the absolute risks are very small in the later time periods.

@@

# vb3064$social_class <- vb3064$hisclass
# vb3064$marital_status <- vb3064$civst
# vb3064$urban_or_rural <- vb3064$urban
# vb3064$age_group <- vb3064$age
fitt <- tpchreg(event ~ offset(log(exposure)) + age + urban + civst + sex * hisclass * period, cuts = 40:65,
             data = vb1Tab)
dr <- drop1(fitt, test = "Chisq")
# ##dr
```

@@

fittold <- coxph(Surv(enter, exit, event) ~ hisclass * period * sex + civst + urban, data = vb2)
drold <- drop1(fittold, test = "Chisq")
```

@@

```{r hazcalw2}
nper <- length(labb)
his <- levels(vb1$hisclass)
nhis <- length(his)
##
resm <- matrix(0, nrow = nhis, ncol = nper)
for (i in 1:nper){
    dat <- vb2[vb2$sex == "male" & vb2$period == labb[i], ]
    dat$wght <- getWeights(dat)
    fit <- coxph(Surv(enter, exit, event) ~ strata(hisclass), weights = wght, data = dat)
    resm[, i] <- summary(survfit(fit), rmean = 100)$table[, "*rmean"]
}
##
resw <- matrix(0, nrow = nhis, ncol = nper)
for (i in 1:nper){
    dat <- vb2[vb2$sex == "female" & vb2$period == labb[i], ]
    dat$wght <- getWeights(dat)
    fit <- coxph(Surv(enter, exit, event) ~ strata(hisclass), weights = wght, data = dat)
    resw[, i] <- summary(survfit(fit), rmean = 100)$table[, "*rmean"]
}

>>>>

```{r readalldata, eval = FALSE}
source("R/get.file.R")
##skume <- get.file("skume")
load("data/skume.rda")
skume <- skume[order(skume$id, skume$enter), ]
skume$id <- paste("pop", skume$id, sep = "_")
##
skume1 <- age.window(skume, c(40, 65))  ###### OJOJOJ!!!!!!!!!!!!!!!!!!!!!!!!!!! (40, 65)
skume1 <- oneSpell(skume1)
##
skume2 <- age.window(skume, c(65, 100))
skume2 <- oneSpell(skume2)
##################################################
##foli <- get.file("foli")
load("data/foli.rda")
foli$civst[is.na(foli$civst)] <- "unmarried"
foli <- foli[order(foli$id, foli$enter), ]
foli$AR <- foli$ULORSAK <- NULL
##
foli1 <- age.window(foli, c(40, 65))
foli1 <- oneSpell(foli1)
##
foli2 <- age.window(foli, c(65, 100))
foli2 <- oneSpell(foli2)
##################################################
source("CreData/R/addPer.R")
skume1 <- addPer(skume1, cuts = seq(1801, 1951, by = 25))
skume2 <- addPer(skume2, cuts = seq(1801, 1951, by = 25))
##
foli1 <- addPer(foli1, cuts = c(1976, 1986, 1996, 2005, 2014))
foli2 <- addPer(foli2, cuts = c(1976, 1986, 1996, 2005, 2014))
##################################################
vb1 <- rbind(skume1, foli1)
vb1$period <- factor(vb1$period)
cuts1 <- seq(40, 65, by = 5) ############################## NOTE!!
vb1Tab <- toTpch(vb1[, !(colnames(vb1) %in% c("id", "birthdate"))], cuts = cuts1)
##
vb2 <- rbind(skume2, foli2)
vb2$period <- factor(vb2$period)
cuts2 <- c(seq(65, 90, by = 5), 100) ###################### NOTE!!
vb2Tab <- toTpch(vb2[, !(colnames(vb2) %in% c("id", "birthdate"))], cuts = cuts2)
vb <- rbind(vb1, vb2)
vb$exposure <- vb$exit - vb$enter # For statistics and figures purposes
vbTab <- rbind(vb1Tab, vb2Tab) # NEW
saveRDS(vbTab, file = "data/vbTab.rds")
saveRDS(vb1Tab, file = "data/vb1Tab.rds")
saveRDS(vb2Tab, file = "data/vb2Tab.rds")
```

```{r nydata}
vbTab <- readRDS("data/vbTab.rds")
vb1Tab <- readRDS("data/vb1Tab.rds")
vb2Tab <- readRDS("data/vb2Tab.rds")
cuts1 <- seq(40, 65, by = 5) ############################## NOTE: fragile!!
cuts2 <- c(seq(65, 90, by = 5), 100) ###################### NOTE: fragile!!
```

@@

```{r hazcalw}
source("R/getWeights.R")
labb <- levels(vbTab$period)
nper <- length(labb)
his <- levels(vb1Tab$hisclass)
nhis <- length(his)
##
resm <- matrix(0, nrow = nhis, ncol = nper)
pdm <-  matrix(0, nrow = nhis, ncol = nper)
for (i in 1:nper){
    dat <- vb1Tab[vb1Tab$sex == "male" & vb1Tab$period == labb[i], ]
    ##dat <- vb1[vb1$sex == "male" & vb1$period == labb[i], ]
    ##dat$wght <- getWeights(dat)
    ##fit <- coxph(Surv(enter, exit, event) ~ strata(hisclass), weights = wght, data = dat)
    ##x <- survfit(fit)
    for (j in seq_along(his)){
        fit <- tpchreg(event ~ offset(log(exposure)) + age, data = dat[dat$hisclass == his[j], ], cuts = seq(40, 65, by = 5))
        haz <- c(0, fit$hazards, fit$hazards[length(fit$hazards)])
        pdm[j, i] <- ppch(65, cuts = fit$cuts, levels = haz)
    }
    ##resm[, i] <- summary(x, rmean = 65)$table[, "*rmean"]
    ##pdm[, i] <- 1 - x$surv[cumsum(x$strata)]
}
##
resw <- matrix(0, nrow = nhis, ncol = nper)
pdw <-  matrix(0, nrow = nhis, ncol = nper)
for (i in 1:nper){
    dat <- vb1Tab[vb1Tab$sex == "female" & vb1Tab$period == labb[i], ]
    ##dat <- vb1[vb1$sex == "female" & vb1$period == labb[i], ]
    ##dat$wght <- getWeights(dat)
    ##fit <- coxph(Surv(enter, exit, event) ~ strata(hisclass), weights = wght, data = dat)
    ##x <- survfit(fit)
    for (j in seq_along(his)){
        fit <- tpchreg(event ~ offset(log(exposure)) + age, data = dat[dat$hisclass == his[j], ], cuts = cuts1)
        haz <- c(0, fit$hazards, fit$hazards[length(fit$hazards)])
        pdw[j, i] <- ppch(65, cuts = fit$cuts, levels = haz)
    }
    ##resw[, i] <- summary(x, rmean = 65)$table[, "*rmean"]
    ##pdw[, i] <- 1 - x$surv[cumsum(x$strata)]
}

@@

Differences in mortality in the age group 40--64 are getting very small as we move forward in time: Almost all forty-year-olds today survive beyond 65, regardless of social class. In Sweden of the year 2014, the probability of surviving age 65 for a forty-year-old man is 91 per cent  (a period measure). The corresponding figure for a forty-year-old woman is 94 per cent. A relevant measure of comparison is the probability (risk) of dying before age 65 for a forty-year-old, remembering that the absolute risks are very small in the later time periods.

@@

# vb3064$social_class <- vb3064$hisclass
# vb3064$marital_status <- vb3064$civst
# vb3064$urban_or_rural <- vb3064$urban
# vb3064$age_group <- vb3064$age
fitt <- tpchreg(event ~ offset(log(exposure)) + age + urban + civst + sex * hisclass * period, cuts = cuts1,
             data = vb1Tab)
dr <- drop1(fitt, test = "Chisq")
# ##dr
```

@@

##fittold <- coxph(Surv(enter, exit, event) ~ hisclass * period * sex ##+ civst + urban, data = vb2)
fittold <- tpchreg(event ~ offset(log(exposure)) + age + hisclass * period * sex + civst + urban, cuts = cuts2, data = vb2Tab)
drold <- drop1(fittold, test = "Chisq")
```

@@

```{r hazcalw2}
nper <- length(labb)
his <- levels(vb2Tab$hisclass)
nhis <- length(his)
##
resm <- matrix(0, nrow = nhis, ncol = nper)
for (i in 1:nper){
    dat <- vb2Tab[vb2Tab$sex == "male" & vb2Tab$period == labb[i], ]
    dat$wght <- getWeights(dat)
    for (j in seq_along(his)){
        fit <- tpchreg(event ~ offset(log(exposure)) + age, 
                       cuts = cuts2, 
                       data = dat[dat$hisclass == his[i], ])
        haz <- c(0, fit$hazards, fit$hazards[length(fit$hazards)])
        cat("lenght(haz) = ", length(haz), "\n")
        resm[j, i] <- mpch(levels = haz, cuts = fit$cuts)
    }
    ##fit <- coxph(Surv(enter, exit, event) ~ strata(hisclass), weights = wght, data = dat)
    ##resm[, i] <- summary(survfit(fit), rmean = 100)$table[, "*rmean"]
}
##
resw <- matrix(0, nrow = nhis, ncol = nper)
for (i in 1:nper){
    dat <- vb2Tab[vb2Tab$sex == "female" & vb2Tab$period == labb[i], ]
    dat$wght <- getWeights(dat)
        for (j in seq_along(his)){
        fit <- tpchreg(event ~ offset(log(exposure)) + age, 
                       cuts = cuts2, 
                       data = dat[dat$hisclass == his[i], ])
        haz <- c(0, fit$hazards, fit$hazards[length(fit$hazards)])
        resw[j, i] <- mpch(levels = haz, cuts = fit$cuts)
    }

    ##fit <- coxph(Surv(enter, exit, event) ~ strata(hisclass), weights = wght, data = dat)
    ##resw[, i] <- summary(survfit(fit), rmean = 100)$table[, "*rmean"]
}
